---
title: "FMT 16S Microbiome Analysis"
author: "April Jauhal"
date: "2026-04-16"
output: html_document
---
This notebook contains code used for figure analysis in figures 1 and 2. In this workflow, 16S data is processed, alpha and beta diversity and ANCOMBC2 analysis is performed, and data is formatted for RPCA analysis (see RPCA.ipynb). Also contains code used to generate Figure 1, as well as example code for the generation of figure 2 (final Figure 2 subfigures were generated in PRISM).  

## Data Processing
Set Directory 
``` {r}
FMT_dir = "C:/Users/ETTINA03/NYU Langone Health Dropbox/April Jauhal/Shenhav_Lab/Morrowitz_16S/2024_Jul_FMT/"
FMT_figpath = paste0(FMT_dir, "/Figures/")
FMT_RPCA_path = paste0(FMT_dir, "/RPCA_3/")
```
Loading libraries
``` {r}
library(dplyr)
library(phyloseq)
library(readxl)
library(stringr)
library(vegan)
library(ggplot2)
library(ggpubr)
library(ANCOMBC)
library(ggprism)
```
Defining colors for plots to align with prism color scheme
``` {r}
five_colors <- c(
  "FMT-ABX" = "#AD07E3",
  "FMT-HD" = "#FF8000",
  "PBS" = "#00C000",
  "Untreated" = "#0000FF", 
  "ABX (2W)" = "#FF0000")
 
three_colors <- c(
  "FMT-ABX" = "#FF0000",
  "FMT-HD" = "#00C000",
  "PBS" = "#0000FF"
)

four_colors <- c(
  "FMT-ABX" = "#AD07E3",
  "FMT-HD" = "#FF8000",
  "PBS" = "#00C000",
  "Untreated" = "#0000FF")
```
Loading 16S dataset and converting to physeq objects
``` {r}
theme_set(theme_bw())

# Create figure path if it doesn't already exist
if (!dir.exists(FMT_figpath)) {
  dir.create(FMT_figpath, recursive = TRUE)
}

count_table = read_xlsx(paste0(FMT_dir, "FMT_July2024_16S_Sequencing_Stats.xlsx"),
                        sheet="Microbial Composition") %>% 
  dplyr::rename(taxonomy=`Amplicon Sequencing Variant (ASV)`) 
sample_info = read_xlsx(paste0(FMT_dir, "FMT_July2024_16S_Sequencing_Stats.xlsx"),
                        sheet="Metadata") %>% 
  as.data.frame() %>% mutate(Cage = paste0(Label, "_", Cage)) %>%
  mutate(Treatment = case_when(Label == "ABX" ~ "ABX (2W)",
                               Label == "Donor" ~ "Donor", 
                               Label == "FMTABX" ~ "FMT-ABX",
                               Label == "FMTCTRL" ~ "FMT-HD",
                               Label == "FMTPBS" ~ "PBS",
                               Label == "Naive" ~ "Untreated",
                               TRUE ~ NA,
                               ))


bridge_otu_df = count_table %>% dplyr::select(`taxonomy`) 
bridge_otu_df$ID = paste0("otu_", 1:nrow(bridge_otu_df))


tax_df <- data.frame(str_split(count_table$taxonomy, ';')) %>% t() %>% data.frame
colnames(tax_df) <- c("Domain", "Phylum",  "Class",  "Order",  "Family",  "Genus", "Species")
tax_df <- tax_df %>% dplyr::mutate(taxonomy = apply(., 1, function(row) paste(row, collapse = ";"))) %>% left_join(bridge_otu_df) %>% dplyr::select(-taxonomy) %>% tibble::column_to_rownames('ID') 
tax_df[] <- lapply(lapply(tax_df, function(x) sub("[a-z]__", "", x)), function(x) sub("^__$", "Unknown", x)) 

new_taxa_table = tax_df %>% slice(0)
for (each_genus in unique(tax_df$Genus)) {
  subset_df = tax_df %>% filter(Genus==each_genus)
  if (each_genus %in% c("uncultured", "Unknown")) {
    subset_df = subset_df %>% mutate(Genus = paste0(Genus, "_", row_number()))
  }
  new_taxa_table = rbind(new_taxa_table, subset_df)
}


otu_df = count_table %>% left_join(bridge_otu_df) %>% select(-taxonomy) %>% tibble::column_to_rownames('ID')
rownames(sample_info) <- sample_info$Sample

# Version for original relative abundance
FMT_physeq_rel = phyloseq(otu_table(otu_df, taxa_are_rows=TRUE), tax_table(as.matrix(new_taxa_table)), sample_data(sample_info))

seq_stats = read_xlsx(paste0(FMT_dir, "FMT_July2024_16S_Sequencing_Stats.xlsx"),
                      sheet="Sequencing Statistics")
count_data = otu_df
for (sampleid in seq_stats$`Sample Name`) {
  num_reads = seq_stats %>% filter(`Sample Name`==sampleid) %>% pull(`Total Read Pairs`)
  count_data[sampleid] <- round(count_table[sampleid]*num_reads, digits = 0)
}

# Version for whole counts 
FMT_physeq_counts = phyloseq(otu_table(as.matrix(count_data), taxa_are_rows=TRUE), tax_table(as.matrix(new_taxa_table)), sample_data(sample_info))
``` 
Filtering out mitochondria and choloroplast DNA and calculating relative abundance
``` {r}
# Version for relative abundance without mitochondria/chloroplasts
FMT_physeq_filtered_rel = transform_sample_counts(subset_taxa(FMT_physeq_rel, !Genus %in% c("Mitochondria", "Chloroplast")), function(x) x / sum(x))

# Version for whole counts without michrondria/chloroplasts
FMT_physeq_filtered_counts = subset_taxa(FMT_physeq_counts, !Genus %in% c("Mitochondria", "Chloroplast"))
```
# Constructing relative abundance barplots (class)
barplots are shown in Figure 1B
``` {r}
library(RColorBrewer)
set.seed(2)
palette <- colorRampPalette(brewer.pal(12, "Paired"))(50)
palette_scrambled <- sample(palette, 50)

# class-level plotting data
plot_data_class0 = FMT_physeq_filtered_rel %>%
  tax_glom(taxrank = "Class") %>%                     
  psmelt() %>% group_by(Class, Sample) %>% dplyr::summarize(class_count = sum(Abundance)) %>% left_join(sample_data(FMT_physeq_filtered_rel)) 
class_1perc = plot_data_class0 %>% group_by(Class) %>% 
  dplyr::summarize(max = max(class_count)) %>%
  filter(max >= 0.01) %>% pull(Class)
plot_data_class = plot_data_class0 %>% mutate(Class = ifelse(Class %in% class_1perc, Class, "Other"))


collapse_class <- function(label, timepoint) {
  df = plot_data_class %>%
      filter(Label==label, Timepoint==timepoint) %>%
            group_by(Label, Class) %>%
      dplyr::summarize(x=sum(class_count), Timepoint=timepoint, Label=label) %>%
        unique() 
  return(df %>% mutate(class_count = x/sum(df$x, na.rm = TRUE)))
}
grouped_plot_data_class <- do.call(rbind,
        list(collapse_class("FMTPBS", 0),
             collapse_class("FMTPBS", 14),
             collapse_class("FMTPBS", 24),
             collapse_class("FMTCTRL", 0),
             collapse_class("FMTCTRL", 14),
             collapse_class("FMTCTRL", 24),
             collapse_class("FMTABX", 0),
             collapse_class("FMTABX", 14),
             collapse_class("FMTABX", 24)))
rel_barplot_class = ggarrange(
  ggplot(grouped_plot_data_class %>%
           mutate(Label = ifelse(Label=="FMTABX", "FMT-ABX",
                                 ifelse(Label=="FMTPBS", "PBS",
                                        ifelse(Label=="FMTCTRL", "FMT-HD", NA)))), aes(y=as.factor(Timepoint), x=class_count, fill=Class)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        legend.position = "bottom", 
        strip.text.y= element_text(size=8),
        strip.text.x= element_text(size=8)) +
  facet_grid(.~Label) +
  scale_fill_manual(values = palette_scrambled) + scale_color_manual(values = palette_scrambled) + ylab("Timepoint") + xlab("Relative Abundance"),
  ggplot(plot_data_class %>% filter(!is.na(FMT_Donor)) %>%
           mutate(Subject = ifelse(Subject=="Donor_Abx", "Antibiotics",
                                 ifelse(Subject=="Donor_CTRL", "Control", NA))), aes(y=as.factor(Subject), x=class_count, fill=Class)) + geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text.y= element_text(size=8),
        strip.text.x= element_text(size=8)) +
  scale_fill_manual(values = palette_scrambled) + scale_color_manual(values = palette_scrambled) + ylab("Donor") + xlab("Relative Abundance"),
  common.legend = TRUE, widths = c(2,1), legend = "bottom")
rel_barplot_class
ggexport(rel_barplot_class, dpi = 600, width = 700, height=200, filename = paste0(FMT_figpath, "rel_barplot_class.png"))
```

## Alpha Diversity analysis
Alpha diversity analysis was initially performed by SeqCenter, results are parsed here and used to generate sublots for Figure 1FG.
``` {r}
Alpha_df = read_xlsx(paste0(FMT_dir, "FMT_July2024_16S_Sequencing_Stats.xlsx"),
                     sheet="Alpha Diversity Metrics")
div_sample_data = sample_data(FMT_physeq_filtered_rel) %>% data.frame() %>% left_join(Alpha_df) 

rownames(div_sample_data) <- div_sample_data$Sample
sample_data(FMT_physeq_filtered_rel) <- sample_data(div_sample_data)

rich_plot = ggplot(div_sample_data %>% filter(!is.na(Timepoint), Timepoint==24) %>%
                     mutate(TP_label = ifelse(Timepoint==24, "Day 24", NA)), 
                   aes(x=Treatment, y=`Observed Features`, color=Treatment)) +
    geom_boxplot(show.legend = FALSE) + geom_jitter(width = 0.2, height = 0, size = 1) +
    facet_grid(.~TP_label) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    theme(legend.position = "left") +
    geom_point() + theme(legend.position = "none") +
    stat_compare_means(comparisons = list(c("FMT-HD", "Untreated"),
                                          c("PBS", "Untreated"),
                                          c("FMT-ABX", "Untreated")), label = "p.signif",
                       bracket.size = 0.75, tip.length = 0) 

faith_plot = ggplot(div_sample_data %>% filter(!is.na(Timepoint), Timepoint==24) %>%
                     mutate(TP_label = ifelse(Timepoint==24, "Day 24", NA)), 
                    aes(x=Treatment, y=`Faith's Phylogenetic Diversity`, color=Treatment)) +
    geom_boxplot(show.legend = FALSE) + geom_jitter(width = 0.2, height = 0, size = 1) +
    facet_grid(.~TP_label) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    theme(legend.position = "left") +
    geom_point() + theme(legend.position = "none") +
    stat_compare_means(comparisons = list(c("FMT-HD", "Untreated"),
                                          c("PBS", "Untreated"),
                                          c("FMT-ABX", "Untreated")), label = "p.signif",
                       bracket.size = 0.75, tip.length = 0) 

rich_prism_plot = rich_plot + theme_prism(base_size = 12) +
  scale_color_manual(values = four_colors) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

faith_prism_plot = faith_plot + theme_prism(base_size = 12) +
  scale_color_manual(values = four_colors) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

alpha_div_plot = ggarrange(rich_prism_plot, faith_prism_plot, nrow = 1)

ggexport(alpha_div_plot,
         dpi = 300, width = 600, height = 400, filename = paste0(FMT_figpath, "diversity.png"))
alpha_div_plot
```

## Beta diversity analysis
Beta diversity calculations were performed on filtered relative abundance data
``` {r}
meta = data.frame(sample_data(FMT_physeq_filtered_rel)) %>%
  mutate(Treatment = case_when(Label == "ABX" ~ "ABX (2W)",
                               Label == "Donor" ~ "Donor", 
                               Label == "FMTABX" ~ "FMT-ABX",
                               Label == "FMTCTRL" ~ "FMT-HD",
                               Label == "FMTPBS" ~ "PBS",
                               Label == "Naive" ~ "Untreated",
                               TRUE ~ NA,
                               ))
otu = t(otu_table(FMT_physeq_filtered_rel))
jsd_res <- distance(t(otu_table(FMT_physeq_filtered_rel)), method='jsd')

jaccard = distance(otu, method='jaccard')

FMT_genus <- tax_glom(FMT_physeq_filtered_rel, taxrank = "Genus", NArm = TRUE) 
otu_genus = t(otu_table(FMT_genus)) 
jaccard_genus = distance(otu_genus, method='jaccard')

FMT_family <- tax_glom(FMT_physeq_filtered_rel, taxrank = "Family", NArm = TRUE) 
otu_family = t(otu_table(FMT_family)) 
jaccard_family = distance(otu_family, method='jaccard')
```
Comparing ABX-treated and FMT-ABX samples to initial timepoint
``` {r}
jaccard_FMT_dur_df = data.frame(Label=character(), Comparison=logical(), TP_Comparison=logical(), Jaccard=numeric())

Sample_list5 = list(meta %>% filter(Label=="FMTCTRL", Timepoint %in% c(24, 14, 0)) %>% pull(Sample),
                    meta %>% filter(Label=="FMTABX", Timepoint %in% c(24, 14, 0 )) %>% pull(Sample),
                    meta %>% filter(Label=="FMTPBS", Timepoint %in% c(24, 14, 0 )) %>% pull(Sample))
for (Samples in Sample_list5) {
  for(each_sample in Samples) {
    
    Sample_Label = meta %>% filter(Sample==each_sample) %>% pull(Treatment)
    Sample_Time = meta %>% filter(Sample==each_sample) %>% pull(Timepoint)
    
    for (compare_sample in unlist(Sample_list5[Sample_list5!=each_sample])) {
      Compare_Sample_Label = meta %>% filter(Sample==compare_sample) %>% pull(Treatment) 
      Compare_Sample_Time = meta %>% filter(Sample==compare_sample) %>% pull(Timepoint) 

      each_df = data.frame(Treatment=Sample_Label, 
                           Comparison = ifelse(Sample_Label==Compare_Sample_Label, 'Self', 'Other'),
                           TP_Comparison =  case_when(Sample_Time == 24 & Compare_Sample_Time == 24  ~ "24 vs. 24 Days",
                                                      Sample_Time == 24 & Compare_Sample_Time == 14 ~ "24 vs. 14 Days",
                                                      Sample_Time == 14 & Compare_Sample_Time == 24 ~ "14 vs. 24 Days",
                                                      Sample_Time == 14 & Compare_Sample_Time == 14 ~ "14 vs. 14 Days",
                                                      Sample_Time == 0 & Compare_Sample_Time == 0 ~ "0 vs. 0 Days",
                                                      Sample_Time == 0 & Compare_Sample_Time == 14 ~ "0 vs. 14 Days",
                                                      Sample_Time == 14 & Compare_Sample_Time == 0 ~ "14 vs. 0 Days",
                                                      Sample_Time == 0 & Compare_Sample_Time == 24 ~ "0 vs. 24 Days",
                                                      Sample_Time == 24 & Compare_Sample_Time == 0 ~ "24 vs. 0 Days",
                                                      TRUE ~ "Other"),
                           Jaccard=as.matrix(jaccard)[each_sample, compare_sample])
      jaccard_FMT_dur_df = rbind(jaccard_FMT_dur_df, each_df)
    }
  }
}
```
Building subplots for figure 1CDE
``` {r}
jaccard_FMPT_dur_subplot1 = jaccard_FMT_dur_df %>% filter(Comparison=="Self", TP_Comparison %in% c('0 vs. 0 Days', '0 vs. 14 Days', '0 vs. 24 Days')) %>% ggplot(aes(x=TP_Comparison, y=Jaccard, color = Treatment)) + geom_boxplot() + 
  geom_jitter(width = 0.2, height = 0, size = 1) + xlab("Comparison") + ylab("Pairwise Jaccard Distances") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    stat_compare_means(comparisons = list(c("0 vs. 0 Days", "0 vs. 14 Days"),
                                          c("0 vs. 14 Days", "0 vs. 24 Days")), 
                       bracket.size = 0.75, tip.length = 0) + facet_grid(.~Treatment)
  
jaccard_FMPT_dur_subplot2 =   jaccard_FMT_dur_df %>% filter(Comparison=="Self", TP_Comparison %in% c('0 vs. 14 Days')) %>% ggplot(aes(x=Treatment, y=Jaccard, color = Treatment)) + geom_boxplot() + geom_jitter(width = 0.2, height = 0, size = 1) +
  xlab("Comparison") + ylab("Pairwise Jaccard Distances") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    stat_compare_means(comparisons = list(c("FMT-ABX", "FMT-HD"),
                                          c("FMT-ABX", "PBS"),
                                          c("PBS", "FMT-HD")), 
                       bracket.size = 0.75, tip.length = 0) + facet_grid(.~TP_Comparison)
jaccard_FMPT_dur_subplot3 =     jaccard_FMT_dur_df %>% filter(Comparison=="Self", TP_Comparison %in% c('0 vs. 24 Days')) %>% ggplot(aes(x=Treatment, y=Jaccard, color = Treatment)) + geom_boxplot() + geom_jitter(width = 0.2, height = 0, size = 1) +
  xlab("Comparison") + ylab("Pairwise Jaccard Distances") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    stat_compare_means(comparisons = list(c("FMT-ABX", "FMT-HD"),
                                          c("FMT-ABX", "PBS"),
                                          c("PBS", "FMT-HD")), 
                       bracket.size = 0.75, tip.length = 0) + facet_grid(.~TP_Comparison)

jaccard_FMPT_dur_subprism1 = jaccard_FMPT_dur_subplot1 + theme_prism(base_size = 12) +
  scale_color_manual(values = three_colors) + theme(legend.position="none",
                                                    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
jaccard_FMPT_dur_subprism2 = jaccard_FMPT_dur_subplot2 + theme_prism(base_size = 12) +
  scale_color_manual(values = three_colors) + theme(legend.position="none",
                                                    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
jaccard_FMPT_dur_subprism3 = jaccard_FMPT_dur_subplot3 + theme_prism(base_size = 12) +
  scale_color_manual(values = three_colors) + theme(legend.position="none",
                                                    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

jaccard_FMT_dur_plot_prism = ggarrange(jaccard_FMPT_dur_subprism1, jaccard_FMPT_dur_subprism2, jaccard_FMPT_dur_subprism3, nrow = 1, widths = c(2,1, 1))
jaccard_FMT_dur_plot_prism

ggexport(jaccard_FMT_dur_plot_prism,
         dpi = 300, width = 600, height = 400, filename = paste0(FMT_figpath, "Treatment_effect.png"))

```
Comparing day 24 samples to their donors for supplementary figure 1
``` {r}
jaccard_donor_df = data.frame(Label=character(), Cagemate=logical(), Jaccard=numeric())

Sample_list2 = list(meta %>% filter(Label=="FMTCTRL", Timepoint==24) %>% pull(Sample),
                    meta %>% filter(Label=="FMTABX", Timepoint==24) %>% pull(Sample),
                    meta %>% filter(Label=="FMTPBS", Timepoint==24) %>% pull(Sample))
for (Samples in Sample_list2) {
  for(each_sample in Samples) {
    
    Sample_Label = meta %>% filter(Sample==each_sample) %>% pull(Treatment)

    for (compare_sample in meta %>% filter(!is.na(FMT_Donor)) %>% pull(Sample)) {
      Donor_type = meta %>% filter(Sample==compare_sample) %>% pull(FMT_Donor) 
      each_df = data.frame(Treatment=Sample_Label, 
                           Donor_type = Donor_type,
                           Jaccard=as.matrix(jaccard)[each_sample, compare_sample])
      jaccard_donor_df = rbind(jaccard_donor_df, each_df)
    }
  }
}

jaccard_donor_plot = jaccard_donor_df %>% 
  mutate(Donor_label = case_when(Donor_type=="Donor_ABX" ~ "ABX-treated Donor",
                                 Donor_type=="Donor_CTRL" ~ "Control Donor",
                                 TRUE ~ NA)) %>%
  ggplot(aes(x=Donor_label, y=Jaccard, color = Treatment)) + geom_boxplot() + geom_jitter(width = 0.2, height = 0, size = 1) + 
  ylab("\nDay 24 Jaccard Distance to Donor") + xlab("Donor Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'none') +
  stat_compare_means(comparisons = list(c("ABX-treated Donor", "Control Donor")),
                     bracket.size = 0.75, tip.length = 0) + facet_grid(.~Treatment) 

jaccard_donor_plot2 = jaccard_donor_df %>% 
  mutate(Donor_label = case_when(Donor_type=="Donor_ABX" ~ "ABX-treated Donor",
                                 Donor_type=="Donor_CTRL" ~ "Control Donor",
                                 TRUE ~ NA)) %>%
  ggplot(aes(x=Treatment, y=Jaccard, color = Treatment)) + geom_boxplot() + geom_jitter(width = 0.2, height = 0, size = 1) + 
  ylab("Day 24 Jaccard Distance to Donor") + xlab("Treatment") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'none') +
  stat_compare_means(comparisons = list(c("FMT-ABX", "FMT-HD"),
                                          c("FMT-ABX", "PBS"),
                                          c("PBS", "FMT-HD")),
                     bracket.size = 0.75, tip.length = 0) + facet_grid(.~Donor_label)

jaccard_donor_prism_plot =jaccard_donor_plot + theme_prism(base_size = 12) +
  scale_color_manual(values = three_colors) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.position = 'none')

jaccard_donor_prism_plot2 = jaccard_donor_plot2 + theme_prism(base_size = 12) +
  scale_color_manual(values = three_colors) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.position = 'none')

ggarrange(jaccard_donor_prism_plot, jaccard_donor_prism_plot2)
ggexport(ggarrange(jaccard_donor_prism_plot, jaccard_donor_prism_plot2),
         dpi = 300, width = 500, height = 400, filename = paste0(FMT_figpath, "Donor_comparison.png"))
```

## Data processing for RPCA analysis
Loaded physeq data is processed for analysis via FMT_RPCA.ipynb
``` {r}
FMT_physeq_nodonor_24 <- subset_samples(subset_samples(FMT_physeq_filtered_counts %>% tax_glom(taxrank = "Genus"), Label %in% c("FMTABX", "FMTCTRL", "FMTPBS")),
                                        Timepoint==24)
sample_metadata_rpca <- data.frame(sample_data(FMT_physeq_nodonor_24)) %>% rename(sampleid=Sample) %>% select(sampleid, everything())
subject_metadata_rpca <- sample_metadata_rpca %>% select(Subject, Label) %>% unique()

# Uncomment to prepare files for CTF
library(biomformat)
FMT_physeq.biom <- make_biom(otu_table(FMT_physeq_nodonor_24))
#write_biom(FMT_physeq.biom, biom_file="paste0(FMT_RPCA_path, "/16S_forRPCA.biom")
# write.table(sample_metadata_rpca, file=paste0(FMT_RPCA_path, "RPCA_sample_metadata.tsv",
#             quote=FALSE, sep='\t', row.names = FALSE)
# write.table(subject_metadata_rpca, file=paste0(FMT_dir, "RPCA_subject_metadata.tsv",
#             quote=FALSE, sep='\t', row.names = FALSE)
```
Loaded results following RPCA analysis via FMT_RPCA.ipynb
``` {r}
ord_text = readLines(paste0(FMT_RPCA_path, "/16S_RPCA/biplot_unzipped/629da861-81c6-48df-a17a-8f24eeca9919/data/ordination.txt"))

feat_ord = read.table(textConnection(ord_text[str_detect(ord_text, "otu")]), header = FALSE, col.names = c("Feature", "PC1", "PC2", "PC3")) %>% 
  left_join(tax_table(FMT_physeq_nodonor_24) %>% 
              as.data.frame() %>% tibble::rownames_to_column("Feature"))
 
otu_filt_counts = FMT_physeq_filtered_counts %>% otu_table %>% as.data.frame()
tax_filt = FMT_physeq_filtered_counts %>% tax_table %>% as.data.frame()
perc_10 = round(nrow(feat_ord)/10) 

PC1_full_num = otu_filt_counts[feat_ord %>% filter(PC1<0) %>% pull(Feature),] %>% colSums()
PC1_full_denom = otu_filt_counts[feat_ord %>% filter(PC1>=0) %>% pull(Feature),] %>% colSums()
PC2_full_num = otu_filt_counts[feat_ord %>% filter(PC2<0) %>% pull(Feature),] %>% colSums()
PC2_full_denom = otu_filt_counts[feat_ord %>% filter(PC2>=0) %>% pull(Feature),] %>% colSums()
PC3_full_num = otu_filt_counts[feat_ord %>% filter(PC3<0) %>% pull(Feature),] %>% colSums()
PC3_full_denom = otu_filt_counts[feat_ord %>% filter(PC3>=0) %>% pull(Feature),] %>% colSums()


PC1_10perc_num = otu_filt_counts[feat_ord %>% arrange(desc(PC1)) %>% slice(1:perc_10) %>% pull(Feature),] %>% colSums()
PC1_10perc_denom = otu_filt_counts[feat_ord %>% arrange(PC1) %>% slice(1:perc_10) %>% pull(Feature),] %>% colSums()
PC2_10perc_num = otu_filt_counts[feat_ord %>% arrange(desc(PC2)) %>% slice(1:perc_10) %>% pull(Feature),] %>% colSums()
PC2_10perc_denom = otu_filt_counts[feat_ord %>% arrange(PC2) %>% slice(1:perc_10) %>% pull(Feature),] %>% colSums()
PC3_10perc_num = otu_filt_counts[feat_ord %>% arrange(desc(PC3)) %>% slice(1:perc_10) %>% pull(Feature),] %>% colSums()
PC3_10perc_denom = otu_filt_counts[feat_ord %>% arrange(PC3) %>% slice(1:perc_10) %>% pull(Feature),] %>% colSums()


PC1_full = log(PC1_full_num/PC1_full_denom) %>% data.frame(PC1_full=.) %>% tibble::rownames_to_column("sampleid")
PC2_full = log(PC2_full_num/PC2_full_denom) %>% data.frame(PC2_full=.) %>% tibble::rownames_to_column("sampleid")
PC3_full = log(PC3_full_num/PC3_full_denom) %>% data.frame(PC3_full=.) %>% tibble::rownames_to_column("sampleid")
PC1_10perc = log(PC1_10perc_num/PC1_10perc_denom) %>% data.frame(PC1_10perc=.) %>% tibble::rownames_to_column("sampleid")
PC2_10perc = log(PC2_10perc_num/PC2_10perc_denom) %>% data.frame(PC2_10perc=.) %>% tibble::rownames_to_column("sampleid")
PC3_10perc = log(PC3_10perc_num/PC3_10perc_denom) %>% data.frame(PC3_10perc=.) %>% tibble::rownames_to_column("sampleid")

log_ratio_df = full_join(full_join(full_join(full_join(full_join(PC1_full, PC2_full),
                                                       PC3_full), 
                                             PC1_10perc), 
                                   PC2_10perc), 
                         PC3_10perc)


subj_ord = read.table(textConnection(ord_text[str_detect(ord_text, "FMT")]), header = FALSE, col.names = c("sampleid", "PC1", "PC2", "PC3")) %>% left_join(log_ratio_df) %>% left_join(sample_metadata_rpca) 

PC2_denom_df = tax_filt[feat_ord %>% arrange((PC2)) %>% slice(1:perc_10) %>% pull(Feature),] %>% dplyr::select(Class, Genus)
PC2_num_df = tax_filt[feat_ord %>% arrange(desc(PC2)) %>% slice(1:perc_10) %>% pull(Feature),] %>% dplyr::select(Class, Genus)
PC3_denom_df = tax_filt[feat_ord %>% arrange((PC3)) %>% slice(1:perc_10) %>% pull(Feature),] %>% dplyr::select(Class, Genus)
PC3_num_df = tax_filt[feat_ord %>% arrange(desc(PC3)) %>% slice(1:perc_10) %>% pull(Feature),] %>% dplyr::select(Class, Genus)

bind_rows(
  feat_ord %>% arrange(desc(PC2)) %>% slice(1:perc_10) %>% mutate(Ranking = paste0("Numerator: ", 1:8)) %>%
    dplyr::select(Ranking, PC2, Class, Genus),
  feat_ord %>% arrange((PC2)) %>% slice(1:perc_10) %>% mutate(Ranking = paste0("Denominator: ", 1:8)) %>%
    dplyr::select(Ranking, PC2, Class, Genus)) %>% write.csv(paste0(FMT_dir, "Table1.csv"))
```
## Displaying RPCA results (Note that the final figure was generated in PRISM)
Figure 2A
``` {r}
Figure2A = ggplot(subj_ord, aes(x = PC2, y = PC3, shape = Label, color = Label)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("FMTABX" = three_colors[["FMT-ABX"]], "FMTCTRL" = three_colors[["FMT-HD"]], "FMTPBS" = three_colors[["PBS"]]),
    labels = c("FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD", "FMTPBS" = "PBS")
  ) +
  scale_shape_manual(
    values = c("FMTABX" = 15, "FMTCTRL" = 17, "FMTPBS" = 16),  # square, triangle, circle
    labels = c("FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD", "FMTPBS" = "PBS")
  ) +
  scale_x_continuous(
    limits = c(-0.6, 0.6),
    breaks = seq(-0.6, 0.6, by = 0.2),
    labels = scales::number_format(accuracy = 0.1)   
  ) +
  scale_y_continuous(
    limits = c(-1.0, 1.0),
    breaks = seq(-1.0, 1.0, by = 0.5)
  ) +
  labs(
    x = "PC2",
    y = "PC3",
    color = NULL,
    shape = NULL
  ) +
  theme_prism(base_size = 12, base_fontface = "bold") +
  theme(
    legend.position    = c(0.85, 0.85),      # top-right inside plot
    legend.background  = element_blank(),
    legend.key         = element_blank(),
    panel.grid         = element_blank()
  )
Figure2A
ggexport(Figure2A, filename = paste0(FMT_figpath, "Figure2A.png"), width=400, height=300)
```
Figure 2B
``` {r}
Figure2B = ggplot(subj_ord, aes(x = Label, y = PC2, fill = Label, color = Label)) +
  geom_boxplot(
    fill  = "white",                         
    width = 0.5,
    outlier.shape = NA                         
  ) +
  geom_point(
    position = position_jitter(width = 0.1),
    size = 2
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = 0.6) +
  stat_compare_means(
    comparisons = list(c("FMTCTRL", "FMTABX"), c("FMTABX", "FMTPBS")),
    label = "p.signif",                       
    bracket.size = 0.75, tip.length = 0,
  ) +
  scale_fill_manual(
    values = c("FMTABX" = three_colors[["FMT-ABX"]], "FMTCTRL" = three_colors[["FMT-HD"]], "FMTPBS" = three_colors[["PBS"]]),
    labels = c("FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD", "FMTPBS" = "PBS"),
    breaks = c("FMTPBS", "FMTABX", "FMTCTRL")
  ) +
  scale_color_manual(
    values = c("FMTABX" = three_colors[["FMT-ABX"]], "FMTCTRL" = three_colors[["FMT-HD"]], "FMTPBS" = three_colors[["PBS"]]),
    labels = c("FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD", "FMTPBS" = "PBS"),
    breaks = c("FMTPBS", "FMTABX", "FMTCTRL")
  ) +
  scale_x_discrete(
    limits = c("FMTPBS", "FMTABX", "FMTCTRL"),   
    labels = c("FMTPBS" = "PBS", "FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD"),
  ) +
  scale_y_continuous(
    limits = c(-0.6, 0.6),
    breaks = seq(-0.6, 0.6, by = 0.2),
    labels = scales::number_format(accuracy = 0.1)   
  ) +
  labs(x = "Label", y = "PC2", fill = NULL, color = NULL) +
  theme_prism(base_size = 12, base_fontface = "bold") +
  theme(
    legend.position   = "none",               
    panel.grid        = element_blank(),
    axis.text.x       = element_text(angle = 45, hjust = 1, vjust = 1)
  )
Figure2B
ggexport(Figure2B, filename = paste0(FMT_figpath, "Figure2B.png"), width=200, height=300)
```
Figure 2C
``` {r}
Figure2C = ggplot(subj_ord, aes(x = Label, y = PC2_10perc, fill = Label, color = Label)) +
  geom_boxplot(
    fill  = "white",
    width = 0.5,
    outlier.shape = NA
  ) +
  geom_point(
    position = position_jitter(width = 0.1),
    size = 2
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", linewidth = 0.6) +
  stat_compare_means(
    comparisons = list(c("FMTCTRL", "FMTABX"),
                       c("FMTCTRL", "FMTPBS"),
                       c("FMTPBS",  "FMTABX")),
    label        = "p.signif",
    bracket.size = 0.75,
    tip.length   = 0
  ) +
  scale_fill_manual(
    values = c("FMTABX" = three_colors[["FMT-ABX"]], "FMTCTRL" = three_colors[["FMT-HD"]], "FMTPBS" = three_colors[["PBS"]]),
    labels = c("FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD", "FMTPBS" = "PBS"),
    breaks = c("FMTPBS", "FMTABX", "FMTCTRL")
  ) +
  scale_color_manual(
    values = c("FMTABX" = three_colors[["FMT-ABX"]], "FMTCTRL" = three_colors[["FMT-HD"]], "FMTPBS" = three_colors[["PBS"]]),
    labels = c("FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD", "FMTPBS" = "PBS"),
    breaks = c("FMTPBS", "FMTABX", "FMTCTRL")
  ) +
  scale_x_discrete(
    limits = c("FMTPBS", "FMTABX", "FMTCTRL"),
    labels = c("FMTPBS" = "PBS", "FMTABX" = "FMT-ABX", "FMTCTRL" = "FMT-HD")
  ) +
  labs(
    x     = "Label",
    y     = "Log(Top PC2 Ranks:Bottom PC2 Ranks)",
    fill  = NULL,
    color = NULL
  ) +
  theme_prism(base_size = 12, base_fontface = "bold") +
  theme(
    legend.position = "none",
    panel.grid      = element_blank(),
    axis.text.x     = element_text(angle = 45, hjust = 1, vjust = 1)
  )
Figure2C
ggexport(Figure2C, filename = paste0(FMT_figpath, "Figure2C.png"), width=200, height=300)
```

## ANCOMBC2 analysis
Main analysis (Results used to construct Figure 2D in PRISM)
``` {r}
set.seed(123)
res_label_24_genus <- ancombc2(data = subset_samples(subset_samples(FMT_physeq_filtered_counts, Label %in% c("FMTABX", "FMTCTRL", "FMTPBS")), Timepoint ==24),
                          tax_level = "Genus",
                          fix_formula = "Label",
                          rand_formula = NULL,
                          p_adj_method = "holm", pseudo_sens = TRUE,
                          prv_cut = 0, lib_cut = 0, s0_perc = 0.05,
                          group = "Label", struc_zero = TRUE, neg_lb = FALSE,
                          alpha = 0.05, n_cl = 2, verbose = TRUE,
                          global = TRUE, pairwise = TRUE, dunnet = TRUE, trend = FALSE,
                          iter_control = list(tol = 1e-2, max_iter = 20, 
                                              verbose = TRUE),
                          em_control = list(tol = 1e-5, max_iter = 100),
                          lme_control = NULL,
                          mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
                          trend_control = NULL)
res_24_genus = res_label_24_genus$res_pair

ggplot(res_24_genus %>% mutate(taxon = factor(res_24_genus$taxon, levels = (res_24_genus %>% arrange(lfc_LabelFMTCTRL) %>% pull(taxon)))),
       aes(y=taxon, x=lfc_LabelFMTCTRL, fill = diff_LabelFMTCTRL)) + geom_bar(stat="identity") + ylab("Genus") + xlab("log fold change FMTCTRL vs. FMTABX") + labs(fill="Significance")

ggplot(res_24_genus %>% mutate(taxon = factor(res_24_genus$taxon, levels = (res_24_genus %>% arrange(lfc_LabelFMTPBS) %>% pull(taxon)))),
       aes(y=taxon, x=lfc_LabelFMTPBS, fill = diff_LabelFMTPBS)) + geom_bar(stat="identity") + ylab("Genus") + xlab("log fold change FMTPBS vs. FMTABX ") + labs(fill="Significance")

ggplot(res_24_genus %>% mutate(taxon = factor(res_24_genus$taxon, levels = (res_24_genus %>% arrange(lfc_LabelFMTPBS_LabelFMTCTRL) %>% pull(taxon)))),
       aes(y=taxon, x=lfc_LabelFMTPBS_LabelFMTCTRL, fill = diff_LabelFMTPBS_LabelFMTCTRL)) + geom_bar(stat="identity") + ylab("Genus") + xlab("log fold change FMTPBS vs. FMTCTRL") + labs(fill="Significance")

res_24_genus %>% mutate(taxon = factor(res_24_genus$taxon, levels = (res_24_genus %>% arrange(lfc_LabelFMTCTRL) %>% pull(taxon)))) %>% dplyr::select(taxon,lfc_LabelFMTCTRL, p_LabelFMTCTRL)

res_24_genus %>% mutate(taxon = factor(res_24_genus$taxon, levels = (res_24_genus %>% arrange(lfc_LabelFMTCTRL) %>% pull(taxon)))) %>% dplyr::select(taxon,lfc_LabelFMTCTRL, p_LabelFMTCTRL, q_LabelFMTCTRL) %>% filter(p_LabelFMTCTRL < 0.05)

res_24_genus %>% 
  mutate(fold_change = exp(lfc_LabelFMTCTRL), log_fold_change = log(fold_change, base = exp(1))) %>%
  dplyr::select(taxon, fold_change, log_fold_change) %>% 
  write.csv(paste0(FMT_dir, "fold_change_values.csv"))

```

``` {r}
# Prepare data with significance stars
res_24_genus_plot_df = res_24_genus %>%
  mutate(
    taxon = factor(taxon, levels = (res_24_genus %>% 
                     arrange(lfc_LabelFMTCTRL) %>% 
                     pull(taxon))),
    # Direction for color (negative = red, positive = green)
    direction = ifelse(lfc_LabelFMTCTRL < 0, "neg", "pos"),
    # Significance determines fill vs outline
    sig = diff_LabelFMTCTRL,
    # Star labels for significant taxa
    stars = case_when(
      q_LabelFMTCTRL < 0.001 ~ "***",
      q_LabelFMTCTRL < 0.01  ~ "**",
      q_LabelFMTCTRL < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  arrange(desc(lfc_LabelFMTCTRL)) 

Figure2C = ggplot(res_24_genus_plot_df, aes(y = taxon, x = lfc_LabelFMTCTRL)) +
  # Outlined (non-significant) bars
  geom_bar(data = filter(res_24_genus_plot_df, !sig),
           stat = "identity",
           fill = "white",
           color = NA,
           alpha = 0.15) +
  # Filled (significant) bars
  geom_bar(data = filter(res_24_genus_plot_df, sig),
           stat = "identity",
           aes(fill = direction),
           color = NA,
           alpha = 1) +
  # Outline for all bars
  geom_bar(stat = "identity",
           aes(color = direction),
           fill = NA,
           linewidth = 0.4) +
  # Vertical line at 0
  geom_vline(xintercept = 0, linewidth = 0.8, color = "black") +
  scale_y_discrete(limits = levels(res_24_genus_plot_df$taxon)) + 
  # Significance stars — placed just beyond bar end
  geom_text(aes(
    x = lfc_LabelFMTCTRL + ifelse(lfc_LabelFMTCTRL >= 0, 0.15, -0.15),
    label = stars,
    hjust = ifelse(lfc_LabelFMTCTRL >= 0, 0, 1)
  ), size = 3.5) +
  # Tick marks at bar ends for significant taxa (mimics prism style)
  geom_segment(data = filter(res_24_genus_plot_df, sig),
               aes(x = lfc_LabelFMTCTRL,
                   xend = lfc_LabelFMTCTRL,
                   y = as.numeric(taxon) - 0.45,
                   yend = as.numeric(taxon) + 0.45),
               color = "black", linewidth = 0.5) +
  # Colors: red for negative, green for positive
  scale_fill_manual(values  = c("neg" = three_colors[["FMT-ABX"]], "pos" = three_colors[["FMT-HD"]])) +
  scale_color_manual(values = c("neg" = three_colors[["FMT-ABX"]], "pos" = three_colors[["FMT-HD"]])) +
  # Prism theme
  theme_prism(base_size = 11, base_fontface = "bold") +
  theme(
    legend.position = "none",
    axis.line.y     = element_blank(),   # remove y-axis line (prism style)
    panel.grid      = element_blank()
  ) +
  labs(
    y = "Genus",
    x = "Log-fold Difference\n(FMTCTRL vs. FMTABX)"
  )
Figure2C
ggexport(Figure2C, filename = paste0(FMT_figpath, "Figure2D.png"), width = 500, height = 500)
```

Additional analysis: To evaluate the hypothesis that the Roseburia enrichment in the FMT-ABX group was already present at day 14, we leveraged ANCOMBC2 analysis for Day 14 data and found that Roseburia was already enriched in FMT-ABX vs. FMT-HD prior to FMT treatment (p=0.032).
``` {r}
genus_meta_df = otu_table(FMT_genus) %>% t() %>% data.frame() %>% tibble::rownames_to_column("Sample") %>%
  left_join(meta)
Roseburia = FMT_physeq_filtered_counts %>% tax_table ()%>% as.data.frame() %>% filter(Genus=="Roseburia") %>% rownames()

ggplot(genus_meta_df %>% filter(Timepoint==14, Label %in% c("FMTABX", "FMTCTRL")), aes(x=Label, y=!!sym(Roseburia))) + geom_boxplot() + facet_grid(.~Timepoint) +
  stat_compare_means(comparisons = list(c("FMTABX", "FMTCTRL"))) + ylab("Roseburia")

```


