---
title: "Misame growth trajectory analysis"
author: "April Jauhal"
date: "2024-04-12"
output: html_document
---
Misame growth trajectory analysis & plotting + additional metadata calculations (6 month growth/length velocities and HCAZ)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Setting to overwrite existing files within document; does not apply to gbmt RDS files which must be deleted manually to overwrite
``` {r}
misame_traj_overwrite=FALSE
```

Loading libraries
``` {r}
library(ggplot2)
library(ggpubr)
library(gbmt)
library(tidyverse)
library(zscorer)
```
Loading data and setting paths 
``` {r}
IMiC_path = "C://Users/ETTINA03/NYU Langone Health Dropbox/April Jauhal/Shenhav_Lab/"
misame_metadata <- read.delim(paste0(IMiC_path, "IMiC/Data/MISAME/metadata/MISAME_3_IMiC_analysis.csv"),
                              sep=",")
traj_RDS_path <- paste0(IMiC_path, "IMiC/Data/MISAME/data/RDS/") # create path before running!
misame_traj_key_path <- paste0(IMiC_path, "IMiC/Data/MISAME/metadata/misame_processed_traj_key.csv")
misame_traj_growth_path <- paste0(IMiC_path, "IMiC/Data/MISAME/metadata/misame_processed_traj_growth.csv")
misame_supp_metadata_path <- paste0(IMiC_path, "IMiC/Data/MISAME/metadata/misame_metadata_supp.csv")
fig_path <- paste0(IMiC_path, "C://Users/ETTINA03/NYU Langone Health Dropbox/April Jauhal/Shenhav_Lab/IMiC/Figures")
```

Extracting growth metric visits from first 6 months
``` {r}
# Excluding recall visits, AGE imputation set to 0 to remove duplicates
misame_no_recall <- subset(subset(misame_metadata, is.na(VISIT_R_FL)), AGEIMPFL==0)
misame_growth_visits <-  c("Delivery", "Post Natal FU M01", "Post Natal FU M02", "Post Natal FU M03", "Post Natal FU M04", "Post Natal FU M05", "Post Natal FU M06")
misame_growth_df <- subset(misame_no_recall, VISIT %in% misame_growth_visits)
```
Defining functions
``` {r}
run_gbmt <- function (ngroups, dim, metric, df) {
  # inputs num groups, dimensions, metric (eg. "WAZ"), and df (likely growth_df)
  # returns gbmt results object, either runs or recovers from RDS file
  # Assumes "traj_RDS_path" defined at the top of the script contains the RDS folders
  # No scaling procedure is run (which makes sense if using z-scores as inputs)
  RDS_name=paste0("m", ngroups, "_", dim, ".RDS")
  WAZ_RDS_path <- paste0(gsub("/$", "", traj_RDS_path), "/", metric, "/")
  rdsfile=paste0(WAZ_RDS_path, RDS_name)
  if (file.exists(dirname(rdsfile))) {
    if (!file.exists(rdsfile)) {
      gbmt_res <- gbmt(x.names=metric, unit="SUBJIDO", time="AGEDAYS", d=dim, ng=ngroups, data=df, scaling=0)
      saveRDS(gbmt_res, rdsfile)
      print("new save file written")
    } else {print("save file exists")}
    return(readRDS(rdsfile))
  } else {print(paste0("ERROR: check/create filepath: ", dirname(file)))}
}

bic_calculate <- function(model_val, df) {
  return(-2*model_val$logLik+model_val$npar*log(nrow(df)))
}

bic_compare <- function(bic_list) {
  # Takes a list input (names=varname, value=value) of BIC values, tests based on < -2 BIC diff threshold
  # Performs pairwise comparisons and returns the single best BIC as a named list
  print("always double check that list components are increasing in complexity")
  diff_vector <- c(names(bic_list)[1])
  for (n in 1:(length(bic_list)-1)) {
    diff <- as.numeric(bic_list[[(n+1)]]) - as.numeric(bic_list[[n]])
    print(paste0(names(bic_list)[(n+1)], "-", names(bic_list)[(n)], ": ", diff))
    if (diff < -2) {
      diff_vector[length(diff_vector)+1] <- names(bic_list)[n+1]
    }
  }
  retval <- list(bic_list[[length(diff_vector)]])
  names(retval) <- names(bic_list)[length(diff_vector)]
  print(paste0("best model: ", names(retval)))
  return(retval)
}
```
# Trajectory calculation for WAZ
Group based trajectory modeling
``` {r Calculating WAZ trajectories}

# Dimension 1
WAZm1_1 <- run_gbmt(ngroups=1, dim=1, metric="WAZ", df=misame_growth_df)
WAZm2_1 <- run_gbmt(ngroups=2, dim=1, metric="WAZ", df=misame_growth_df)
WAZm3_1 <- run_gbmt(ngroups=3, dim=1, metric="WAZ", df=misame_growth_df)
WAZm4_1 <- run_gbmt(ngroups=4, dim=1, metric="WAZ", df=misame_growth_df)
WAZm5_1 <- run_gbmt(ngroups=5, dim=1, metric="WAZ", df=misame_growth_df)
WAZm6_1 <- run_gbmt(ngroups=6, dim=1, metric="WAZ", df=misame_growth_df)

# Dimension 2
WAZm1_2 <- run_gbmt(ngroups=1, dim=2, metric="WAZ", df=misame_growth_df)
WAZm2_2 <- run_gbmt(ngroups=2, dim=2, metric="WAZ", df=misame_growth_df)
WAZm3_2 <- run_gbmt(ngroups=3, dim=2, metric="WAZ", df=misame_growth_df)
WAZm4_2 <- run_gbmt(ngroups=4, dim=2, metric="WAZ", df=misame_growth_df)
WAZm5_2 <- run_gbmt(ngroups=5, dim=2, metric="WAZ", df=misame_growth_df)
WAZm6_2 <- run_gbmt(ngroups=6, dim=2, metric="WAZ", df=misame_growth_df)

# Dimension 3
WAZm1_3 <- run_gbmt(ngroups=1, dim=3, metric="WAZ", df=misame_growth_df)
WAZm2_3 <- run_gbmt(ngroups=2, dim=3, metric="WAZ", df=misame_growth_df)
WAZm3_3 <- run_gbmt(ngroups=3, dim=3, metric="WAZ", df=misame_growth_df)
WAZm4_3 <- run_gbmt(ngroups=4, dim=3, metric="WAZ", df=misame_growth_df)
WAZm5_3 <- run_gbmt(ngroups=5, dim=3, metric="WAZ", df=misame_growth_df)
WAZm6_3 <- run_gbmt(ngroups=6, dim=3, metric="WAZ", df=misame_growth_df)

# Dimension 4
WAZm1_4 <- run_gbmt(ngroups=1, dim=4, metric="WAZ", df=misame_growth_df)
WAZm2_4 <- run_gbmt(ngroups=2, dim=4, metric="WAZ", df=misame_growth_df)
WAZm3_4 <- run_gbmt(ngroups=3, dim=4, metric="WAZ", df=misame_growth_df)
WAZm4_4 <- run_gbmt(ngroups=4, dim=4, metric="WAZ", df=misame_growth_df)
WAZm5_4 <- run_gbmt(ngroups=5, dim=4, metric="WAZ", df=misame_growth_df)
WAZm6_4 <- run_gbmt(ngroups=6, dim=4, metric="WAZ", df=misame_growth_df)

# Dimension 5
WAZm1_5 <- run_gbmt(ngroups=1, dim=5, metric="WAZ", df=misame_growth_df)
WAZm2_5 <- run_gbmt(ngroups=2, dim=5, metric="WAZ", df=misame_growth_df)
WAZm3_5 <- run_gbmt(ngroups=3, dim=5, metric="WAZ", df=misame_growth_df)
WAZm4_5 <- run_gbmt(ngroups=4, dim=5, metric="WAZ", df=misame_growth_df)
WAZm5_5 <- run_gbmt(ngroups=5, dim=5, metric="WAZ", df=misame_growth_df)
WAZm6_5 <- run_gbmt(ngroups=6, dim=5, metric="WAZ", df=misame_growth_df)

```
Comparing 1 dimensional models
``` {r}
WAZ_dim1_best <- bic_compare(list(WAZ_BIC_1_1=bic_calculate(WAZm1_1, misame_growth_df), 
                                  WAZ_BIC_2_1=bic_calculate(WAZm2_1, misame_growth_df), 
                                  WAZ_BIC_3_1=bic_calculate(WAZm3_1, misame_growth_df), 
                                  WAZ_BIC_4_1=bic_calculate(WAZm4_1, misame_growth_df), 
                                  WAZ_BIC_5_1=bic_calculate(WAZm5_1, misame_growth_df), 
                                  WAZ_BIC_6_1=bic_calculate(WAZm6_1, misame_growth_df)))

print(WAZ_dim1_best)
```
Comparing 2 dimensional models
```{r}
WAZ_dim2_best <- bic_compare(list(WAZ_BIC_1_2=bic_calculate(WAZm1_2, misame_growth_df), 
                                  WAZ_BIC_2_2=bic_calculate(WAZm2_2, misame_growth_df), 
                                  WAZ_BIC_3_2=bic_calculate(WAZm3_2, misame_growth_df), 
                                  WAZ_BIC_4_2=bic_calculate(WAZm4_2, misame_growth_df), 
                                  WAZ_BIC_5_2=bic_calculate(WAZm5_2, misame_growth_df), 
                                  WAZ_BIC_6_2=bic_calculate(WAZm6_2, misame_growth_df)))
print(WAZ_dim2_best)
```
Comparing 3 dimensional models
```{r}
WAZ_dim3_best <- bic_compare(list(WAZ_BIC_1_3=bic_calculate(WAZm1_3, misame_growth_df), 
                                  WAZ_BIC_2_3=bic_calculate(WAZm2_3, misame_growth_df), 
                                  WAZ_BIC_3_3=bic_calculate(WAZm3_3, misame_growth_df), 
                                  WAZ_BIC_4_3=bic_calculate(WAZm4_3, misame_growth_df), 
                                  WAZ_BIC_5_3=bic_calculate(WAZm5_3, misame_growth_df), 
                                  WAZ_BIC_6_3=bic_calculate(WAZm6_3, misame_growth_df)))
print(WAZ_dim3_best) 
```
Comparing 4 dimensional models
``` {r}
WAZ_dim4_best <- bic_compare(list(WAZ_BIC_1_4=bic_calculate(WAZm1_4, misame_growth_df), 
                                  WAZ_BIC_2_4=bic_calculate(WAZm2_4, misame_growth_df), 
                                  WAZ_BIC_3_4=bic_calculate(WAZm3_4, misame_growth_df), 
                                  WAZ_BIC_4_4=bic_calculate(WAZm4_4, misame_growth_df), 
                                  WAZ_BIC_5_4=bic_calculate(WAZm5_4, misame_growth_df), 
                                  WAZ_BIC_6_4=bic_calculate(WAZm6_4, misame_growth_df)))
print(WAZ_dim4_best) 
```
Comparing 5 dimensional models
``` {r}
WAZ_dim5_best <- bic_compare(list(WAZ_BIC_1_5=bic_calculate(WAZm1_5, misame_growth_df), 
                                  WAZ_BIC_2_5=bic_calculate(WAZm2_5, misame_growth_df), 
                                  WAZ_BIC_3_5=bic_calculate(WAZm3_5, misame_growth_df), 
                                  WAZ_BIC_4_5=bic_calculate(WAZm4_5, misame_growth_df), 
                                  WAZ_BIC_5_5=bic_calculate(WAZm5_5, misame_growth_df), 
                                  WAZ_BIC_6_5=bic_calculate(WAZm6_5, misame_growth_df)))
print(WAZ_dim5_best) 
```
Comparing best models from each dimension
``` {r}
bic_compare(c(WAZ_dim1_best, WAZ_dim2_best, WAZ_dim3_best, WAZ_dim4_best, WAZ_dim5_best))

# The 5-dimensional model provided the most biologically coherent stratification of longitudinal growth trajectories
WAZ_best_model <- WAZ_dim5_best
WAZ_group_list <- get(str_replace(names(WAZ_best_model), "_BIC_", "m"))$assign.list

WAZ_group_assignment <- vector()
for (each_infant in misame_growth_df$SUBJIDO){
  for (WAZ_group_num in names(WAZ_group_list)) {
    if (each_infant %in% WAZ_group_list[[WAZ_group_num]]) {
      WAZ_group_assignment <- c(WAZ_group_assignment, WAZ_group_num)
    }
  }
}
misame_growth_df$WAZ_traj <- as.factor(WAZ_group_assignment)
```
Adding labels
``` {r}
misame_growth_df2 = misame_growth_df %>%
         filter(SUBJID!=1828) %>%  # outlier infant (WAZ values below -6 and continuously trending down)
         mutate(Month=AGEDAYS/30, 
                WAZ_traj_cat = ifelse(WAZ_traj==1, "Stable3",
                                      ifelse(WAZ_traj==2, "Stable2",
                                             ifelse(WAZ_traj==3, "Recovered",
                                                    ifelse(WAZ_traj==4, "Stable4",
                                                           ifelse(WAZ_traj==5, "Unrecovered", "Stable1"))))),
                WAZ_traj_cat = factor(WAZ_traj_cat, levels = c("Stable1", "Stable2", "Stable3", "Stable4", "Recovered", "Unrecovered")))
```

Save results as a key for future analysis 
``` {r}
misame_traj_key = misame_growth_df2 %>% dplyr::rename(SubjectID=SUBJID) %>% dplyr::select(SubjectID, WAZ_traj, WAZ_traj_cat) %>% unique() 
if (misame_traj_overwrite | !file.exists(misame_traj_key_path)) {
  write.csv(misame_traj_key, file=misame_traj_key_path)
}

if (misame_traj_overwrite | !file.exists(misame_traj_growth_path)) {
  write.csv(misame_growth_df2, file=misame_traj_growth_path)
}
```
# Additional metadata calcuations
Building supplementary metadata
``` {r}
weight_diff_df <- left_join(misame_metadata, inner_join(
  misame_metadata %>% filter(VISIT=="Delivery", is.na(VISIT_R_FL)) %>% dplyr::select(SUBJID, WTKG) %>% dplyr::rename(wt_0=WTKG),
  misame_metadata %>% filter(VISIT=="Post Natal FU M06", is.na(VISIT_R_FL)) %>% dplyr::select(SUBJID, WTKG, AGEDAYS) %>% dplyr::rename(wt_6=WTKG, AGE=AGEDAYS))) %>%
    mutate(delta_weight_6mo = wt_6-wt_0,
           growth_vel_6mo=(1000*log(wt_6/wt_0))/AGE) # Patel exponential method

length_diff_df <- left_join(misame_metadata, inner_join(
  misame_metadata %>% filter(VISIT=="Delivery", is.na(VISIT_R_FL)) %>% dplyr::select(SUBJID, LENCM) %>% dplyr::rename(len_0=LENCM),
  misame_metadata %>% filter(VISIT=="Post Natal FU M06", is.na(VISIT_R_FL)) %>% dplyr::select(SUBJID, LENCM, AGEDAYS) %>% dplyr::rename(len_6=LENCM, AGE=AGEDAYS))) %>%
    mutate(delta_length_6mo = len_6-len_0,   
           length_vel_6mo=delta_length_6mo/(AGE/30.437))

hcaz_df = addWGSR(misame_metadata, sex= "SEXN", firstPart = "HCIRCM", secondPart = "AGEDAYS", index = "hca") %>% filter(VISIT=="Post Natal FU M04", is.na(VISIT_R_FL)) %>% dplyr::select(SUBJIDO, hcaz) %>% dplyr::rename(HCAZ_M04 = hcaz) %>% unique()

if (misame_traj_overwrite | !file.exists(misame_supp_metadata_path)) {
  write.csv(weight_diff_df %>% select(SUBJIDO, growth_vel_6mo) %>% full_join(length_diff_df %>% select(SUBJIDO, length_vel_6mo)) %>% full_join(hcaz_df) %>% unique(), misame_supp_metadata_path)
}
```

